Accelerating parameter inference with graphics processing units

### 2019-03-21 paper/referee/response-v0.tex

\begin{document}

We appreciate the referee's thoughtful suggestions, which will improve the paper.

\begin{enumerate}
\item * The introduction should be expanded to (briefly) discuss
hardware-based acceleration techniques in GW data analysis. For
example, there are significant efforts on the search side to use GPUs.
Within parameter estimation communities, much effort has gone into
optimizing LALInference including the use of multi-core environments
(using OpenMP, for example).

\noindent \textbf{Response}:  We added a sentence about pycbc, a search/parameter inference code which makes extensive
use of GP, and another sentence highlighting code optimizations in lalinference.

We regret that a more extended discussion of non-GPU changes to how lalinference's parallelization is implemented (e.g., MPI in the original
lalinference paper vs OpenMP)  would distract the reader,  and that a suitable citation to that project specifically
does not seem available (after consulting with two of the developers responsible).

@@

\noindent \textbf{Response}: While the memory footprint of our operations loosely grows linearly with the number of
modes involved, we have noticed no significant limitations on our memory footprint.  We have colleagues who use models with modes
$l \le 8$ without reporting memory footprint problems.

The memory footprint of the initially transferred data is dominated by the $Q_{k,lm}$ time arrays is no larger than $16 kHz
\times 0.15  \lesssim 2500$ floating point numbers per mode per detector.  Even with several hundred modes, these are
quite small.
Potentially more problematic are our intermediate data products, notably the matrix of time-shifted arrays $Q$ built
from $Q_{k,lm}(\lambda,\tau)$ .  Our code has been carefully
organized to loop over detectors so the memory footprint of $Q$ is only (modes)$\times$ (extrinsic)$\times$ (time),
which (because we perform evaluations in batches of  $10^4$ extrinsic parameters at a time) is roughly $10^4$ times the
footprint of all the underlying $Q_{k,lm}(\lambda,\tau)$ timeseries.  So many modes can
fit well within a typical 4Gb GPU memory space, even allowing for substantial overhead.

\item * What is epsilon appearing in 2c? Is it possible to quote typical
values? Is it problem-specific?

\noindent \textbf{Response}:   The factor $\epsilon$ connecting the number of effective samples to the number of Monte
Carlo trials is indeed highly problem dependent, and reflects the convergence rate of the Monte Carlo integral.
Approximately equivalently, the stanard deviation of the Monte Carlo error estimate scales as $1/\sqrt{N_{it}}$, with a
prefactor  that  depends significantly on the specific integrand $f$ (e.g., $\sigma^2= \frac{1}{N}\int (f^2 - \E{f}^2) dx$)

### 2019-03-21 paper/referee/response-v0.tex

\noindent \textbf{Response}: While the memory footprint of our operations loosely grows linearly with the number of
modes involved, we have noticed no significant limitations on our memory footprint.  We have colleagues who use models with modes
$l \le 8$ without reporting memory footprint problems.

The memory footprint of the initially transferred data is dominated by the $Q_{k,lm}$ time arrays is no larger than $16 kHz
\times 0.15  \lesssim 2500$ floating point numbers per mode per detector.  Even with several hundred modes, these are
quite small.
Potentially more problematic are our intermediate data products, notably the matrix of time-shifted arrays $Q$ built
from $Q_{k,lm}(\lambda,\tau)$ .  Our code has been carefully
organized to loop over detectors so the memory footprint of $Q$ is only (modes)$\times$ (extrinsic)$\times$ (time),
which (because we perform evaluations in batches of  $10^4$ extrinsic parameters at a time) is roughly $10^4$ times the
footprint of all the underlying $Q_{k,lm}(\lambda,\tau)$ timeseries.  So many modes can
fit well within a typical 4Gb GPU memory space, even allowing for substantial overhead.

\item * What is epsilon appearing in 2c? Is it possible to quote typical
values? Is it problem-specific?

\noindent \textbf{Response}:   The factor $\epsilon$ connecting the number of effective samples to the number of Monte
Carlo trials is indeed highly problem dependent, and reflects the convergence rate of the Monte Carlo integral.
Approximately equivalently, the stanard deviation of the Monte Carlo error estimate scales as $1/\sqrt{N_{it}}$, with a
prefactor  that  depends significantly on the specific integrand $f$ (e.g., $\sigma^2= \frac{1}{N}\int (f^2 - \E{f}^2) p_s(x)dx$)

@@

\item * So the reader can more easily compare the performance diagnostics,
please state the most relevant hardware specs of the GPU/CPUs used in
Sec 3 (processor speed, number of cores on the GPU, and memory size)

\noindent \textbf{Response}:   We add an Appendix to describe the hardware used in our study.

@@

\noindent \textbf{Response}: We modified the text to add the word ``approximately'' before uniformly.

In practice, our grid layout code returns precisely the number of points requested.  It attempts to choose the sizes of
each edge of the hypercube to best conform to but be strictly larger than the target size, and then truncates the set of
candidate points.

\item * To give the timing numbers (ie the walltime) a bit more context,
please consider adding the time it would take to do a similar analysis
(same recovery model, set of modes, and similar CPU hardware) with
LALInference. Even rough estimates would be very helpful towards
understanding the full impact of the GPU-accelerated rapidPE code. I
feel like this is one of the most important points, as it is the key
to the impact of this work.

\noindent \textbf{Response}:  In the conclusions, we  have added anecdotal  timing
information on an analysis of practical interest.

That said, this anecdote does not employ identical settings in both codes.  A direct apples-to-apples comparison between RIFT and LI using \emph{identical} waveforms and settings to
what we benchmark here can be grossly unfair to LI.  LI developers
have highly prioritized the use of fast approximations, and reduced-order-quadratures that further accelerate the use of
these approximations.     To the extent these approximations don't impact results, LI can often perform comparable
calculations much more quickly than a LI configuration which used apples-to-apples settings.

@@

\noindent \textbf{Response}:  \editremark{fix me: rcparams}

\item * I would appreciate it if the authors could explain to me why the
evidence converges much more quickly in figure 3 as compared to
figures 1 and 2. Or said in a different way, way are the shapes of
evidence vs iteration so different. They may optionally wish to
comment on this in the paper too.

\noindent \textbf{Response}:  The convergence rate of the evidence depends on the system studied and, critically, the
initial grid: if we have a very well-chosen initial starting grid, the fit  converges rapidly, potentially after just
one iteration.  The configuration used
in Fig 3 has a different dimensionality (4d, not 2d) and a larger starting  grid than our lower-dimensionality investigation.

Qualitatively speaking and based on our large but anecdotoally-built experience, we anticipate that  due to strong degeneracy in
spin -- only one spin degree of freedom is well-constrained -- the seemingly-sparse grid in ${\cal M}_c$ and $\delta$ is very
informative because several redundant configurations with similar $\chi_{\rm eff}$ but different $\chi_{i,z}$ inform our
fit.

### 2019-03-21 paper/paper.tex
referee: response editing
paper: more on response, providing profiling anecdotes as requested

% BORROWED FROM ROM: edit
Several strategies have been
developed to reduce the 
%overall computational cost and turnaround time of 
computational cost of parameter estimation \cite{gwastro-pe-Brandon-STF2,gwastro-PE-AlternativeArchitectures,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,2016PhRvD..94d4031S,2017CQGra..34k5006V,2018arXiv180608792Z}.  
% The computational cost of parameter estimation can be reduced 
Approaches that have appeared in the literature include
%by 
generating the approximate solutions more quickly \cite{gwastro-mergers-PE-ReducedOrder-2013,2014PhRvX...4c1006F,2013PhRvD..87l2002S,2013PhRvD..87d4008C,gwastro-mergers-IMRPhenomP,gwastro-SpinTaylorF2-2013}; 
%by 
interpolating some combination
of the waveform or likelihood \cite{gwastro-approx-ROMNR-Blackman2015,2013PhRvD..87d4008C,2013PhRvD..87l2002S,2014CQGra..31s5010P,2014PhRvD..90d4074S,gwastro-PE-AlternativeArchitectures,cole2014likelihood,2012MNRAS.421..169G};  or 
%by 
adopting a sparse representation to reduce the computational cost of data
handling \cite{antil2013two,gwastro-mergers-PE-ReducedOrder-2013,2016PhRvD..94d4031S,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,gwastro-PE-AlternativeArchitectures,2010arXiv1007.4820C,2018arXiv180608792Z}.   % 2014PhRvX...4c1006F,2013PhRvD..87l2002S,2013PhRvD..87d4008C
 Some methods, however, achieve rapid turnaround through simplifying approximations.  
%
The RIFT/rapidPE strategy described in \cite{gwastro-PE-AlternativeArchitectures,2017CQGra..34n4002O,gwastro-PENR-RIFT,NRPaper} eschews these simplifications, performing
embarrassingly-parallel inferences even for costly models.   However, the method as previously implemented still had a
significant net  computational cost and noticeable startup time, limiting its diverse potential applications.
%
Previously, B. Miller  developed a custom-coded  implementation of the
relevant likelihood on graphics processing units (GPUs), suggesting substantial performance improvements were possible
\cite{Thesis-Brandon-2016,GPU-NU-work}.  
\AddedResponse{The pycbc search code\cite{2016CQGra..33u5004U} and its pycbc-inference extension
  \cite{2019PASP..131b4503B} can also make use of several GPU
  hardware-accelerated operations.
}
In this paper,  we describe a variant of the RIFT approach which flexibly translates one of its
algorithms to operate on GPUs, and adjusts its workflow to exploit the speed improvements
this re-implementation affords.

@@

We illustrate the code using two synthetic signals: a binary black hole with masses $m_{1,2}=35,30 M_\odot$ and a binary
neutron star with masses $1.4 M_\odot$ and $1.35 M_\odot$.
 We perform this analysis on heterogeneous LVC collaboration computing resources, to assess our variable
performance across architectures.  Notably, we investigate the following GPU options: (a) GTX 1050 Ti at LIGO-CIT,
available in large numbers; (b) V100, available on selected high-performance machines; (c) and
GTX 1050 at LIGO-LHO.    Unless otherwise noted in the text, we will discuss code configurations using CPU-only and GPU (b) using
$4096\unit{Hz}$ sampling and only the  $\ell=2, m=\pm 2$ modes.  In this section, we  conservatively report computational cost for
parameters $\bm{\lambda}$ which are close to or within the posterior.  Because of their consistency with the data, 
the marginalized likelihood calculations converge the most slowly, as they have the best-determined extrinsic
parameters.

@@

@@

>>>>

% BORROWED FROM ROM: edit
Several strategies have been
developed to reduce the 
%overall computational cost and turnaround time of 
computational cost of parameter estimation \cite{gwastro-pe-Brandon-STF2,gwastro-PE-AlternativeArchitectures,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,2016PhRvD..94d4031S,2017CQGra..34k5006V,2018arXiv180608792Z}.  
% The computational cost of parameter estimation can be reduced 
Approaches that have appeared in the literature include
%by 
generating the approximate solutions more quickly \cite{gwastro-mergers-PE-ReducedOrder-2013,2014PhRvX...4c1006F,2013PhRvD..87l2002S,2013PhRvD..87d4008C,gwastro-mergers-IMRPhenomP,gwastro-SpinTaylorF2-2013}; 
%by 
interpolating some combination
of the waveform or likelihood \cite{gwastro-approx-ROMNR-Blackman2015,2013PhRvD..87d4008C,2013PhRvD..87l2002S,2014CQGra..31s5010P,2014PhRvD..90d4074S,gwastro-PE-AlternativeArchitectures,cole2014likelihood,2012MNRAS.421..169G};  or 
%by 
adopting a sparse representation to reduce the computational cost of data
handling \cite{antil2013two,gwastro-mergers-PE-ReducedOrder-2013,2016PhRvD..94d4031S,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,gwastro-PE-AlternativeArchitectures,2010arXiv1007.4820C,2018arXiv180608792Z}.   % 2014PhRvX...4c1006F,2013PhRvD..87l2002S,2013PhRvD..87d4008C
\AddedResponse{Not published but also important are code optimization projects that improve    infrastructure,
  such as better parallelization for \texttt{lalinference} \cite{gw-astro-PE-lalinference-v1,lalsuite}.}
 Some methods, however, achieve rapid turnaround through simplifying approximations.  
%
The RIFT/rapidPE strategy described in \cite{gwastro-PE-AlternativeArchitectures,2017CQGra..34n4002O,gwastro-PENR-RIFT,NRPaper} eschews these simplifications, performing
embarrassingly-parallel inferences even for costly models.   However, the method as previously implemented still had a
significant net  computational cost and noticeable startup time, limiting its diverse potential applications.
%
Previously, B. Miller  developed a custom-coded  implementation of the
relevant likelihood on graphics processing units (GPUs), suggesting substantial performance improvements were possible
\cite{Thesis-Brandon-2016,GPU-NU-work}.  
\AddedResponse{The pycbc search code\cite{2016CQGra..33u5004U} and its pycbc-inference extension
  \cite{2019PASP..131b4503B} can also make use of several GPU
  hardware-accelerated operations.
}
In this paper,  we describe a variant of the RIFT approach which flexibly translates one of its
algorithms to operate on GPUs, and adjusts its workflow to exploit the speed improvements
this re-implementation affords.

@@

We illustrate the code using two synthetic signals: a binary black hole with masses $m_{1,2}=35,30 M_\odot$ and a binary
neutron star with masses $1.4 M_\odot$ and $1.35 M_\odot$.
 We perform this analysis on heterogeneous LVC collaboration computing resources \AddedResponse{described in more detail
   in the Appendix}, to assess our variable
performance across architectures.  Notably, we investigate the following GPU options: (a) GTX 1050 Ti at LIGO-CIT,
available in large numbers; (b) V100, available on selected high-performance machines; (c) and
GTX 1050 at LIGO-LHO.    Unless otherwise noted in the text, we will discuss code configurations using CPU-only and GPU (b) using
$4096\unit{Hz}$ sampling and only the  $\ell=2, m=\pm 2$ modes.  In this section, we  conservatively report computational cost for
parameters $\bm{\lambda}$ which are close to or within the posterior.  Because of their consistency with the data, 
the marginalized likelihood calculations converge the most slowly, as they have the best-determined extrinsic
parameters.

@@

\AddedResponse{In addition to producing results rapidly, RIFT results can be produced with a noticably smaller overall resource footprint than
  loosely similar LI analyses, even without tuning to optimize  RIFT pipeline settings.  As a concrete and non-optimized
  example,  for all five of the iterations of the spinning binary neutron star parameter inference with \texttt{SEOBNRv4\_ROM} from
  $20\unit{Hz}$ described in this work, our  RIFT analysis expended  roughly 14 core-days.  By contrast, a TaylorF2 analysis of GW170817
  with LI in MCMC mode starting from $23\unit{Hz}$ required 228 core-days.  % C02-1_TF2LS
  For precessing binary black holes using SEOBNRv3, the improvement is equally substantial: 10 core-days for a
  10-iteration investigation of a synthetic GW150914-like source with RIFT, versus 291 core-days for a LI analysis of
  GW170729.  
  We defer detailed relative benchmarking using comparably-converged parameter inference to future work.
}

@@

\appendix
\section{Properties of resources used}
% https://ldas-gridmon.ligo.caltech.edu/condor/
%
% > CIT:
%% > 560 nodes with single GTX 1050 Ti
%% > 32 nodes with dual K10 PCI cards (shows up as 4 GPU devices) and dual GTX 750 Ti
%% > 16 nodes with dual GTX 750 Ti (and dual Xeon Phi co-processor)
%% > 
%% > LHO and LLO each:
%% > 40 nodes with GTX 1050 Ti
%% > single DGX-1 system with 8 tightly coupled V100

\AddedResponse{
LIGO-CIT worker nodes, denoted by (a)  in the text, are principally S6 nodes with 2-CPU $\times 8$ core Opteron
2.3 GHz machines with 16 Gb of RAM, with GTX 1050 Ti cards with 4 Gb of RAM.  For LIGO-CIT, profiling reports reflect
performance averaged over the whole cluster and hence lacks the detailed reporting produced for the other configurations.
% 
LIGO-LHO worker nodes with GPUs, denoted by (c) in the text, are a heterogeneous configuration mostly consisting of
(a).  Unlike profiling at LIGO-CIT, the profiling for LIGO-LHO reflects controlled tests on a single node.
% nvidia-smi
The V100 machine (ldas-pcdev13, denoted by (b) in the text) is  a 24-core ES-2650 v4 machine with 4 GPUs, only one of
which is active in our tests: Tesla V100 with 16 Gb of RAM.
All non-GPU profiling was  also performed on this machine.  
}
